--- 
title: "Stat 850: Computing Tools for Statisticians"
author: "Susan Vanderplas"
date: "`r paste('Last Updated: ', Sys.Date())`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Introductions to statistical computing packages and document preparation software. Topics include: graphical techniques, data management, Monte Carlo simulation, dynamic document preparation, presentation software."
---


# Introduction {-}

```{r setup, echo = F}
library(magrittr)
library(knitr)
knitr::opts_chunk$set(collapse = T, cache = T)
```

```{r load-libraries, echo = F, message = F, warning = F, cache = F}

saspath <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en"
sasopts <- "-ls 75"
knitr::opts_chunk$set(
  engine.path = list(sas = saspath, saslog = saspath, sashtml = saspath),
  engine.opts = list(sas = sasopts, saslog = sasopts, sashtml = sasopts),
  comment = NA, cache = T
)
library(SASmarkdown)

```

```{r book-functions, include = F}
.pkgkeeplist <- c("package:stats", "package:graphics", "package:grDevices", "package:utils", "package:datasets", "package:methods", "package:base", "package:SASmarkdown", "package:rmarkdown", "package:knitr", "package:magrittr")

.objkeeplist <- c(".pkgkeeplist", ".objkeeplist", ".reset", "saspath", "sasopts")

.reset <- function() {
  basic.packages <- .pkgkeeplist

  package.list <- search()[ifelse(unlist(gregexpr("package:", search())) == 1, TRUE, FALSE)]

  package.list <- setdiff(package.list, basic.packages)

  obj.list <- setdiff(ls(), .objkeeplist)

  if (length(package.list) > 0) for (package in package.list) detach(package, character.only = TRUE)
  rm(list = obj.list)
}
```

```{r pkg-list-update, include = F, cache = F}
# update package list
source("code/collect_packages.R")
```

```{r fa-setup, include=FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```
## Forward {-}

This textbook is intended to be a substitute for hours and hours of video lectures wherein I read code to you. That's awful, and boring^[If you want boring, or have insomnia, check out [this Microsoft Word tutorial from 1989](https://www.youtube.com/watch?v=Jk71bPz5VLo). Youtube says it's the most boring video ever made.], and honestly, as painful for me to have to listen to myself while editing the things as it is for you to watch them.

Instead, I hope  that you'll be able to work through this book, week by week, over the course of Stat 850. I've included comics, snark, gifs, YouTube videos, and more, with the goal of making this a collection of the best information I could find on learning R and SAS and statistical programming. 


## Course Description {-}
Introductions to statistical computing packages and document preparation software. Topics include: graphical techniques, data management, Monte Carlo simulation, dynamic document preparation, presentation software.

## Course Goals {-}
(Broad, amorphous conceptual things that are hard to measure)

1. Learn how to use R and SAS for data analysis, data processing, and data visualization.

2. Become familiar with the process, techniques, and goals of exploratory data analysis.

3. Create, assess, and debug code effectively.

    i. Use online resources to find software to perform a task, comparing approaches taken by competing programs.
    
    ii. Read error messages, find related problems in online forums, and isolate the conditions necessary to generate the error.
    
    iii. Generate minimum working examples or reproducible examples of errors in order to ask for help effectively.

4. Communicate statistical results using reproducible, dynamic tools. Understand the importance of reproducibility in scientific computation.

## Course Objectives {-}
(what you should be able to do at the end of this course)

A. Clean and format the data appropriately for the intended analysis or visualization method. (Goals: 1)

B. Explore a data set using numerical and visual summaries, developing questions which can be answered using statistics. (Goals: 1, 2)

C. Evaluate methods or software to assess relevance to a problem. Compare similar options to determine which are more appropriate for a given application (Goals: 1, 3)

D. Test and debug software, using the following sequence: (Goals: 3, 4)

   1. Reproduce the error in a new environment,
  
   2. Create a minimal reproducible example,
  
   3. Research the error message and evaluate online resources for relevance,
   
   4. Ask for help, describing the error or problem appropriately.

E. Document the data, methods, and results of an analysis using reproducible methods. (Goals: 1, 2, 4)
