---
title: "Homework: Packages, Scripts, Documents, and Functions"
author: "Firstname Lastname"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3)
```

```{r load-sas-libraries, echo = F, message = F, warning = F}

saspath <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en" # Your SAS path goes here
sasopts <- "-ls 75"
knitr::opts_chunk$set(engine.path=list(sas=saspath, saslog=saspath, sashtml=saspath),
                      engine.opts=list(sas=sasopts, saslog=sasopts, sashtml=sasopts), 
                      comment=NA)
library(SASmarkdown)
```

## Instructions
This homework has two main problems. In the FizzBuzz section, you should only modify the code chunks. In the Packages/Procedures section, you should modify the text (anything that doesn't have a `###` in front of it) and the code chunks. Feel free to add additional code chunks as necessary.

Make sure you also edit the author name at the top of the document. 

When you are finished with the assignment: 

1. Save the file as 03_Packages, Scripts, Documents, and Functions/03_Lastname_Firstname.Rmd (filling in the assignment details) and compile it
2. Commit the Rmd file and corresponding html file to your homework git repo
3. Push the commit to github
4. Locate the Rmd file on github and click on the SHA link for the file.
5. Paste the url of the resulting page into Canvas to submit your assignment.

Your assignment must compile on a different computer as long as the saspath is set correctly for that machine.

## FizzBuzz Functions

The "FizzBuzz Test" is a famous programming interview question. 
> Write a program that prints the numbers from 1 to 30. But for multiples of three print “Fizz” instead of the number and for the multiples of five print “Buzz”. For numbers which are multiples of both three and five print “FizzBuzz

In the R chunk below, write a function which will solve this problem. Do not change the function name or parameter name. 

Show the results from running your function on the integers 1-20, as well as a sequence of 10 random integers between 0 and 100. (This code is already written for you).

```{r fizzbuzz-solution, echo = T}

fizzbuzz <- function(x) {
  # FizzBuzz code goes here
}

```

```{r test-fizzbuzz}
fizzbuzz(1:20)
fizzbuzz(sample(1:100, 10)) # sample 10 values from 1 to 100

```

In the sas chunk, write similar code which will solve this problem.

```{sashtml fizzbuzz-sas, error = T}
proc printto log="saslog.log" new;
proc iml;

start fizzbuzz(x);

/* FizzBuzz code goes here */

  
finish;

rndint = rand("Integer", 0, 100);
call fizzbuzz(rand);

x1 = repeat(20, 0, 1); /* initialize x1 */
fb1 = repeat(20, 0, 1); /* initialize fb1 */
  
do i = 1 to 20;
  x1[i] = i;
  fb1[i] = call fizzbuzz(i);
end;

x2 = repeat(10, 0, 1); /* initialize x2 */
fb2 = repeat(10, 0, 1); /* initialize fb2 */
do i = 1 to 10;
  x2[i] = rand("Integer", 1, 100);
  fb2[i] = fizzbuzz(x2[i]);
end;

print x1, fb1;
print x2, fb2;

quit;
```

```{r readlog}
cat(readLines("saslog.log"), sep="\n")
```


```{r cleanuplog}
# Do not forget to remove the log file when you are done!
unlink("saslog.log")
```


## Exploring Packages and Modules

### R Package: XXX
Find an R package which is interesting to you. Fill in the package name above, and describe or demonstrate some feature of the package here. 

### SAS Procedure: XXX
Look through the [SAS Procedures Guide](https://go.documentation.sas.com/?docsetId=proc&docsetTarget=part-2.htm&docsetVersion=9.4&locale=en) (or [this](https://go.documentation.sas.com/?docsetId=proc&docsetTarget=p0ptdedyvyouspn1erbeqr0floca.htm&docsetVersion=9.4&locale=en) guide to procedures documented elsewhere) and find a procedure that is interesting to you that we haven't yet discussed (which is most of them). Read the documentation of that procedure and demonstrate its use on a data set that is appropriate. Do not just copy the example code - actually explore what the procedure does. 

Fill in your code in the chunk below.
```{sashtml proc-demo, error = T}
proc printto log="saslog.log" new;


````


```{r readlog2}
cat(readLines("saslog.log"), sep="\n")
```


```{r cleanuplog2}
# Do not forget to remove the log file when you are done!
unlink("saslog.log")
```

